Characterizing Stellar Streams by ML

### 2023-08-29 src/tex/ms.tex
[showyourwork] overleaf sync

        \subsubsection{On-Stream} \label{sub:method:photometric_model:on_stream}

@@

        In \autoref{sub:method:likelihood_setup} we introduced the likelihood
        setup. In \autoref{sub:method:photometric_model:on_stream} we added
        priors in the form of the isochrone's PDMF and photometric observational
        constraints. In this section we introduce and develop more priors that
        are important for training the BMDN's parameters' neural network(s).

@@

        The custom model framework combines for each model a neural network with
        Bayesian probability functions. The latter are set by the model type,
        e.g. the Gaussian model has a Gaussian likelihood function, but the
        former, the neural network, is set and determined by the user. So long
        as the network has the correct number of inputs (e.g. one for $\phi_1$)
        and outputs (one per parameter), any feed-forward network may be used.
        Our default choice, and for which we provide a helper function, is an
        MLP with a sequence of \texttt{Linear} \!\!\textrightarrow \texttt{Tanh}
        \!\!\textrightarrow \texttt{Dropout} blocks. This is illustrated in
        \autoref{table:mlp_structure}.
        % FIXME! reference not being picked up correctly.
        Dropout layers are used to prevent network co-adaptation and
        over-fitting. Dropout works by randomly zeroing out nodes in the
        network, decreasing the importance of any single node, encouraging
        instead emergent patterns across the network.  One model for which the
        lin-tanh-dropout network cannot be used is the normalizing flow used in
        the photometric background model
        (\autoref{sub:method:photometric_model:off_stream}). Here we use the
        \package{nflows} package, with a conditional diagonal normal base
        distribution and composite transform network of a reverse permutation
        and masked affine auto-regressive transform layers. As noted in our data
        availability statement, all the codes used in this paper are bundled by
        \href{https://github.com/showyourwork/showyourwork}{\package{showyourwork}}
        in a public repository at $\GitHubURL{}$. We will describe for each
        stream model specifics relevant to that stream but refer readers to the
        publicly-available code itself for code implementation details.
        % TODO: Maybe more details in the Appendix?

@@

    We select on-stream and off-stream regions using the known $\phi_1, \phi_2$
    range of the mock stream. The upper panel of
    \autoref{fig:mock_data_photometric_background_selection} shows these
    selections blue and black, respectively. The lower panel shows the on and
    off-stream selections in photometric coordinates. In this space it is
    apparent that the on-stream region has two populations, drawn from the
    isochrone and from background points included in the simple $\phi_1, \phi_2$
    box. In contrast the off-stream region, has only the the background population.
    We use the off-stream population to train a conditional normalizing flow
    (see \autoref{sub:method:photometric_model:off_stream}) on the magnitudes
    $g$ and $r$ and conditioned on $\phi_1$. The normalizing flow is now a
    non-parametric distribution characterizing the photometric background,
    including its variation over $\phi_1$. For this mock stream the flexibility
    of the normalizing flow greatly exceeds necessity: the background is
    drawn from a $\phi_1$-invariant multivariate normal, essentially the base
    distribution of the normalizing flow. It is therefore no surprise that
    the trained normalizing flow provides an excellent characterization of
    the background. We will later see that the flow performs similarly well
    on complex distributions. Having trained the background photometric model
    we proceed to train the full model.

>>>>

        \subsubsection{On-Stream} \label{ssub:method:photometric_model:on_stream}

@@

        In \autoref{sub:method:likelihood_setup} we introduced the likelihood
        setup. In \autoref{ssub:method:photometric_model:on_stream} we added
        priors in the form of the isochrone's PDMF and photometric observational
        constraints. In this section we introduce and develop more priors that
        are important for training the BMDN's parameters' neural network(s).

@@

        The custom model framework combines for each model a neural network with
        Bayesian probability functions. The latter are set by the model type,
        e.g. the Gaussian model has a Gaussian likelihood function, but the
        former, the neural network, is set and determined by the user. So long
        as the network has the correct number of inputs (e.g. one for $\phi_1$)
        and outputs (one per parameter), any feed-forward network may be used.
        Our default choice, and for which we provide a helper function, is an
        MLP with a sequence of \texttt{Linear} \!\!\textrightarrow \texttt{Tanh}
        \!\!\textrightarrow \texttt{Dropout} (hereafter \texttt{LinTanh}) blocks. This is illustrated in
        \autoref{table:mlp_structure}.
        % FIXME! reference not being picked up correctly.
        Dropout layers are used to prevent network co-adaptation and
        over-fitting. Dropout works by randomly zeroing out nodes in the
        network, decreasing the importance of any single node, encouraging
        instead emergent patterns across the network.  One model for which the
        lin-tanh-dropout network cannot be used is the normalizing flow used in
        the photometric background model
        (\autoref{sub:method:photometric_model:off_stream}). Here we use the
        \package{nflows} package, with a conditional diagonal normal base
        distribution and composite transform network of a reverse permutation
        and masked affine auto-regressive transform layers. As noted in our data
        availability statement, all the codes used in this paper are bundled by
        \href{https://github.com/showyourwork/showyourwork}{\package{showyourwork}}
        in a public repository at $\GitHubURL{}$. We will describe for each
        stream model specifics relevant to that stream but refer readers to the
        publicly-available code itself for code implementation details.
        % TODO: Maybe more details in the Appendix?

@@

    We select on-stream and off-stream regions using the known $\phi_1, \phi_2$
    range of the mock stream.  The upper panel of
    \autoref{fig:mock_data_photometric_background_selection} shows these
    selections blue and black, respectively.  The lower panel shows the on and
    off-stream selections in photometric coordinates.  In this space it is
    apparent that the on-stream region has two populations, drawn from the
    isochrone and from background points included in the simple $\phi_1, \phi_2$
    box.  In contrast the off-stream region, has only the the background population.
    We use the off-stream population to train a conditional normalizing flow
    (see \autoref{sub:method:photometric_model:off_stream}) on the magnitudes
    $g$ and $r$ and conditioned on $\phi_1$.  The normalizing flow is now a
    non-parametric distribution characterizing the photometric background,
    including its variation over $\phi_1$.  For this mock stream the flexibility
    of the normalizing flow greatly exceeds necessity: the background is
    drawn from a $\phi_1$-invariant multivariate normal, essentially the base
    distribution of the normalizing flow.  It is therefore no surprise that
    the trained normalizing flow provides an excellent characterization of
    the background.  We will later see that the flow performs similarly well
    on complex distributions.  Having trained the background photometric model
    we proceed to build and train the full model.

    Per \autoref{eq:stream_and_bkg_prob} the stream is a conditionally independent
    addition of both astrometric and photometric models.
    In astrometrics the stream is a Gaussian track in $\phi_2$ and
    parallax $\varpi$, using the model defined in
    \autoref{ssub:method:astrometric_model:on_stream}. We use a 4-layer 
    (\texttt{LinTanh} with shape (1 input, 64 hidden, 4 output), where the 4 outputs
    are $[\mu_{\phi_2}, \sigma_{\phi_2}, \mu_{\varpi}, \sigma_{\varpi}](\phi_1)$.
    The network is small, but sufficiently large to permit 15\% dropout and retain
    robust predictions.
    The photometric model is a distance modulus-shifted isochrone, using the model
    outlined in  \autoref{ssub:method:photometric_model:on_stream}. We use the
    same isochrone parameters as were used to simulate the stream. For this mock
    stream we demonstrate simultaneous astrometric and photometric modeling, linking
    the distance modulus and parallax track, as explained in
    \autoref{ssub:method:linking_to_astrometrics}. Thus the photometric model
    does not have its own network, using instead the outputs
    $[\mu_{\varpi}, \sigma_{\varpi}](\phi_1)$.
    The background model is split in two components, the first is a uniform distribution
    in $\phi_2$, the second an exponential.
